---
title: "Spatiotemporal arrays"
author: "Edzer Pebesma"
date: "1/25/2021"
output: html_document
---

Questions regarding the three papers:

* what is ARD?
* what is the difference between ARD, data cubes, and image collections?
* must ARD, or a data cube, have a regular time discretisation?
* What is a _coverage_, as defined e.g. by OGC in the web coverage service?
* is an image collection always a coverage?
* what does the _stack_ method do in the BDC, for temporal aggregation? 
* why, in the Maus et al. paper, do the authors propose _time-weighted_ dynamic time warping, as opposed to regular dynamic time warping?
* which other questions came up when you read the paper?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###  R's matrix() and image()

```{r}
m = matrix(1:12, 3, 4)
m
image(m)
```

```{r}
fortunes::fortune("conventions")
```

```{r}
image2 = function(x, ...) {
    x = t(x) # transpose
    x = x[,ncol(x):1] # flip
    image(x, ...)
}
image2(m)
```

```{r}
n = 25; m = 20
a = array(runif(n * m), c(n, m))
dim(a)
image2(a, asp = 1)
```

```{r}
fw = 5 # filter width
a1 = apply(a, 1, filter, filter = rep(1, fw)/fw)
image2(a1, asp = 1)
```
We applied the filter to the rows of the matrix, but see the correlation appear in the columns of the resulting matrix. This is because `apply` puts the dimension over which we iterated (rows) last, so rows have become colums:
```{r}
dim(a1)
```

We can further process this two get filtering (moving averaging) on the second dimension, which is again rows of `a1`:
```{r}
a2 = apply(a1, 1, filter, filter = rep(1, fw)/fw)
image2(a2, asp = 1)
dim(a2)
```

We see that the border two rows and columns are all NA:
```{r}
a2[1:5,1:5]
```

We can also apply functions that erase the dimension, e.g. mean:
```{r}
apply(a2, 1, mean, na.rm = TRUE) # row means
apply(a2, 2, mean, na.rm = TRUE) # column means
mean(a2, na.rm = TRUE)
```

### package `stars`
```{r}
library(stars)
m = matrix(1:12, 3, 4)
m.st = st_as_stars(m)
plot(m.st)
```
```{r}
str(m.st)
class(m.st)
is.list(m.st) # is it a list?
length(m.st) # how many elemenst has this list?
class(m.st[[1]]) # what's in its first element?
dim(m.st[[1]])
m.st[[1]]
```

### more dimensions

```{r}
m = array(1:60, c(3,4,5))
m
m.st = st_as_stars(m)
m.st
plot(m.st, text_values = TRUE, text_color = 'orange')
```
### higher dimensional data:
```{r}
dims = c(3,4,5,6,7)
m = array(seq_len(prod(dims)), dims)
m.st = st_as_stars(m)
m.st
```


### real data

```{r}
tif = system.file("tif/L7_ETMs.tif", package = "stars") # a Landsat-7 image part inside the package
r = read_stars(tif)
r
plot(r)
```

Note that:
* `offset`, `delta` and `refsys` allow moving from array index to world coordinates, back and forth
* the third dimension here refers to spectral bands, this can also be time, or third time + fourth band etc.
* `point` registers (_can_ register) whether cell values refer to points or areas (support)
* `values` may give values (e.g. band names, or continuous values if discretisation of the dimensions is _not_ regular)
* `[x]` and `[y]` indicate which variables map the horizontal (hopefully: x and y) dimensions
* a "discrete axis" can refer to a set of (unordered) geometries, leading to vector data cubes

Example of a time series, with curvilinear coordinates:

```{r}
library(dplyr)
prec_file = system.file("nc/test_stageiv_xyt.nc", package = "stars")
(prec = read_ncdf(prec_file, curvilinear = c("lon", "lat"), ignore_bounds = TRUE))
sf::read_sf(system.file("gpkg/nc.gpkg", package = "sf"), "nc.gpkg") %>%
  st_transform(st_crs(prec)) -> nc # transform from NAD27 to WGS84
nc_outline = st_union(st_geometry(nc))
plot_hook = function() plot(nc_outline, border = 'red', add = TRUE)
prec %>%
  slice(index = 1:12, along = "time") %>%
  plot(downsample = c(5, 5, 1), hook = plot_hook)
```