# what is a data cube?

* what is an `vector`?

```{r}
(v = runif(10))
is.vector(v)
```
  
* what is an `array`?

```{r}
a = array(1:24, c(2, 3, 4))
is.vector(a)
dim(a)
a
```

# What is beneficial in arrays?

* why don't we store rgb images in a database with records having fields (row, col, color, intensity)?
* why don't we store videos in a database with records having fields (row, col, frame, color, intensity)?
* what makes an array different from a database with a set of such records?

# Why "cube"?

* comes from OLAP cube: https://en.wikipedia.org/wiki/OLAP_cube , where time and space are (implicit) present as dimensions!
* number of dimensions can be larger than 3
* can the number of dimensions be 2, or 1?
* number of spatial dimensions: 1 (vector data cube), or 2 or 3 (raster data cube), more
* is a `data.frame` a data cube?
* can a `data.frame` hold a data cube?

# is a spatial raster (layer) a data cube?

* what are its necesary properties, compared to a `matrix`?
* can a `data.frame` hold a raster layer?

# did we use data cubes in this course, so far?

* if so, when?

# what are typical cube operations?

* slice (drop 1 or more dimensions)
* cut / crop (create a sub-cube)
* apply some function to the cube
* reduce dimensions (e.g. compute an index from all bands, or a time series summary)

# what is an image collection?

# how can we create a data cube from an image collection?

* deal with varying CRS
* sync time: resample, or aggregate?
* `gdalcubes` https://cran.r-project.org/web/packages/gdalcubes/index.html
    * dimensions: x / y / time / band
    * define extent, target resolution for each dimension
	* define resampling / aggregation strategy 
	* optionally define a reducer function while creating the cube

# what about vector data cubes?

* less familiar, less thought of as "cubes"
* no less useful!
