# Analysis of spatio-temporal data, WS 2020/2021

# Tour the table:

please share:

* who you are, where you are from, and what kind of BSc program you did
* your prior experience with data analysis: what kind of data, what kind of analysis
* your expectations and hopes for this class

# Goals of this course

For the participants to develop

* an idea about which types of spatio-temporal data there are, what kind of questions we can ask to these data, and how these data can be analysed (what kind of models do we use, what kind of diagnostic tools do we use to assess models or assumptions)
* hands-on experience with some of these analysis methods
* experience with common difficulties found in real world datasets
* software skills for analysing spatio-temporal data
* skills to make analyses reproducible, and communicate these

# Introduction

1. What are spatiotemporal data? (is it spatio-temporal or spatiotemporal?)
1. What is involved in data analysis? (exploratory, confirmatory, predictive)
1. How do we analyse data?
1. What is data science? Is geoinformatics a data science?
1. What is [R](https://www.r-project.org/)? 
1. Why do we use 
    - R?
    - [R-markdown](http://rmarkdown.rstudio.com/)?
	- [rstudio](http://rstudio.com/)?

# 2. Exercise Oct 14:

For those *unfamiliar* with R (note that this is a lot of work!):

1. Start R
1. Open http://www.r-project.org/ in a browser
1. Browse to Documentation: Manuals
1. Open: An Introduction to R (base R)
1. Work your way through chapters 1-9. (base R)
1. Alternative: [ModernDive: getting started with R](https://moderndive.com/1-getting-started.html) (tidyverse)
1. Alternative: [R for data science](http://r4ds.had.co.nz/) (tidyverse)


Some general advice:

* always work with the latest released verion of R and Rstudio, and of packages
* if on Mac OSX or Windows, use binary packages unless you know what a compiler is
* regularly update your packages with `update.packages()`
* if installation fails, try to remove a package first with `remove.package()`
* never ignore warnings: they are there for a purpose
* feel free to use other software environments or languages, but make sure assignments are reproducible.

For the following data:

```{r eval=FALSE}
x = c(1,2,4,5)
y = c(2,1,5,4)
data.frame(x,y)
plot(x,y)
```

## Review/refresher questions:

1. How is the mean for a random variable $X$ defined?
1. Give the equation for covariance between two variables $x$ and $y$
1. Give the equation for correlation between two variables $x$ and $y$
1. For the samples $x$ and $y$, compute the sample mean, the variance and the standard deviation 
1. Draw the data in a scatter plot (by hand, or using R)
1. From the data above, compute the covariance between $x$ and $y$, and the correlation between $x$ and $y$
1. Compute a linear regression of $y$ as a function of $x$

For a real dataset, `environmental` from package `lattice` repeat the above questions, and use R. Use variable
`ozone` as `y`, and `temperature` as `x`.

```{r}
data(environmental, package = "lattice")
```
also plot the output of `lm()`, and interpret the resulting diagnostic plots.

