# Time series: frequency domain

Frequency domain time series looks at recurring elements. These are e.g.
caused by:
* astronomical processes
  * sun: day-night: temperature, tides
  * sun: yearly: phenological, plant physiology
  * moon: tidal variations
* social
  * week, months (?)
* biological: 
  * heartbeat, blood pressure
* vibration:
  * two flowing media: waves (wind-water), ripples (wind-sand, water-sand)
  * musical instruments
* other
  * anything that turns, or walks

To review: Trigonometry, e.g. http://en.wikipedia.org/wiki/Trigonometry

The angle transformation formula is given here:

$$\sin(A \pm B) = \sin A \cos B \pm \cos A \sin B$$

In case of a function of time $t$, shifted with a phase shift $\phi$:

$$ f(t) = sin(t + \phi) = \sin t \cos \phi + \cos t \sin \phi = a \sin t + b \cos t$$

with $a = \cos \phi$ and $b = \sin \phi$

Try the following code:
```{r eval=FALSE}
x = seq(0, 4*pi, length.out = 200) # x plotting range

f = function(dir, x) { # plot the combination of a sin+cos function, based on dir
	a = sin(dir)
	b = cos(dir)
	# three curves:
	plot(a * sin(x) + b * cos(x) ~ x, type = 'l', asp=1, col = 'green')
	lines(x, a * sin(x), col = 'red')
	lines(x, b * cos(x), col = 'blue')
	# legend:
	lines(c(10, 10+a), c(2,2), col = 'red')
	lines(c(10, 10), c(2,2+b), col = 'blue')
	arrows(x0 = 10, x1 = 10+a, y0 = 2, y1 = 2+b, .1, col = 'green')
	title("red: sin(x), blue: cos(x), green: sum")
}

for (d in seq(0, 2*pi, length.out = 100)) {
	f(d, x)
	Sys.sleep(.1) # give us some time to think this over!
}
```

## Local regression: LOWESS, LOESS

Moving (unweighted) average filters, shown in
```{r}
set.seed(13531)
n = 200
x = (1:n)/n * 2 * pi
y = sin(x) + 0.5 * rnorm(n)
plot(x, y)
lines(x, filter(y, rep(1/3, 3)), col='red')
lines(x, filter(y, rep(1/10, 10)), col='blue')
lines(x, filter(y, rep(1/30, 30)), col='green')
```

generate the smooth pattern, but are themselves not smooth: the
nearest $n$ observations taken is always a _discrete_ process. When
we move the filter in a continuous way (along $x$), new points
suddenly enter the window, and obtain a positive weight ($1/n$). This
causes the moving average to be _discontinuous_.

A continuous, _smooth_ function is obtained by assigning non-constant
weights such that

* observations, at the edge of entering/exiting the window, have a weight of zero
* (as a consequence:) observations closer to the middle of the window have maximum weight

We can construct such a filter, e.g. by
```{r}
weights = 1 - seq(-1, 1, length.out = 20)^2
weights = weights / sum(weights) # sum to 1
plot(weights, type = 'b')
plot(x, y)
lines(x, filter(y, weights), col='blue', lwd = 2)
```

Local regression (LOWESS, LOESS) is an improvement over weighted moving average
filters, in that it does not only allow a weighted mean, but also a weighted
regression model; in the following plot the lines are shifted vertically by 0.05 and
0.1, because otherwise they would entirely overplot each other:
```{r}
plot(x, y)
lines(x, predict(loess(y ~ x)), col = 'red') # local linear trend
lines(x, .05 + predict(loess(y ~ x + x^2)), col = 'green') # local quadratic trend
lines(x, .10 + predict(loess(y ~ 0 * x)), col = 'orange') # mean
```

We can manipulate the width of the moving window by parameter `span`, e.g.
```{r}
plot(x, y)
lines(x, predict(loess(y ~ x)), col = 'red') # default span: .75
lines(x, predict(loess(y ~ x, span = 0.1)), col = 'green') # narrow window
```

## How to fit a loess to a periodic effect?

_In case we know the period of a periodic effect_, we can transform
the $x$ variable to an angle, by $$2 \pi (x \mod a) * 2 \pi$$ with $a$ the period and
mod the remainder of the division (modulo). Next, windows should be computed in angular
space, meaning that $2 \pi$ equals 0.

## stl, HoltWinters

stl and HoltWinters are two approaches to fit models to time series that contain a
periodic component and a trend:

```{r}
plot(stl(co2, 12))
hw = HoltWinters(co2)
plot(fitted(hw))
plot(hw)
```

Do read the documentation of both functions, and try to understand what they do.
