# Factor and character data

First of all, it is important to realize that R often reads character data as `factor` (categorical) variables. E.g. after using `read.csv`, you might see a variable that looks like time, e.g. in a data.frame:
```{r echo=FALSE}
df = data.frame(month = c("2001-01", "2001-02"), value = c(33.5, 22.9))
```
```{r}
df
```
where it turns out the `month` values are `factor` objects:
```{r}
class(df$month)
```
In such cases, you may have to convert them to character first, before attempting any of the manipulations shown below:
```{r}
df$month = as.character(df$month)
df
class(df$month)
```
Note that they are _not printed differently_, when printing the data.frame. When printing the column only, or `summary(df)`, you will see differences.

# Time and time series data in R

## Date and DateTime 

Date or DateTime values in R can be numbers (`numeric`), or can have one of the
two built-in date / datetime classes. In the first case, we could for instance
have decimal year:
```{r}
(year = c(1995, 1995.25, 1995.5, 1995.75, 1996, 1996.25))
```
denoting a regular series with quarterly time stamps expressed as decimal year.
Nothin about this data reveals that these numbers refer to time stamps, or periods.

### Date

Date values are integers with a class, 
```{r}
s = Sys.Date()
as.integer(s) # nr of days since 1970-01-01
class(s)
```
which print as strings which we can recognize as dates:
```{r}
s
```
We can also read Date from text:
```{r}
as.Date(c("2008-12-31", "2009-01-02"))
```
but only if formatted like YYYY-MM-DD or YYYY/MM/DD. In case of other formats, we need to specify `format`:
```{r}
as.Date(c("2008-31-12", "2009-02-01"), format = "%Y-%d-%m")
```

### POSIXct
Similarly, DateTime objects, 
```{r}
s = Sys.time()
as.numeric(s) # seconds since Jan 1, 1970, midnight UTC
class(s)
```
print like human-readable time stamps:
```{r}
s
```
These can be read from strings formed like "YYYY-mm-dd hh:MM:SS" and some varieties,
```{r}
as.POSIXct("2005-12-31 20:31:55")
```
but in other cases one has to specify `format` e.g. like
```{r}
as.POSIXct("2005-31-12 20.31.55", format = "%Y-%d-%m %H.%M.%S")
```
help is found in `?strptime`, capitalization is important.

### Others: Monthly and Quarterly data

Quarterly data, e.g. encoded as
```{r}
qua = c("2001Q1", "2001Q2", "2001Q3", "2001Q4", "2002Q1")
```
or monthly data encoded like
```{r}
mo = c("2001-01", "2001-02", "2001-03")
```
is  obviously not numeric. We can convert it e.g. into decimal year, e.g. by
```{r}
(qua.ny = as.numeric(
 sub("Q4", ".75",
  sub("Q3", ".5",
   sub("Q2", ".25",
    sub("Q1", ".00", qua))))))
```
for months:
```{r}
(mo.ny = as.numeric(substr(mo, 1, 4)) + (as.numeric(substr(mo, 6, 7)) - 1)/12)
```
(alternatively, given that there are no gaps one can convet such data directly into a `ts` object; see below)

or we can convert it into Date, by
```{r}
(qua.d = as.Date(
 sub("Q4", "-10-01",
  sub("Q3", "-07-01",
   sub("Q2", "-04-01",
    sub("Q1", "-01-01", qua))))))
(mo.d = as.Date(paste0(mo, "-01")))
```

Note that we do different things here:
```{r}
diff(qua.ny)
diff(qua.d)
diff(mo.ny)
diff(mo.d)
```
Depending on the application, one of the two might be the "better" approach:

* if we want to do time series analysis that needs regular time series (time domain analysis, spectral analysis) we want the regular one
* if we are looking at time-extensive variables, e.g. the amount of export or emmissions over a period, taking the difference in period lengths into account _first_ may prevent us from finding spurious periodicities, caused by differences in period (month, quarter) durations

## time series objects

We see that a dataset like
```{r}
head(faithful)
```
is void of any reference to time, whereas 
```{r}
head(co2)
```
does have time reference. The latter is a time series object: R knows that the measurements refer to time steps.

### ts

The simples, and (only?) built-in (i.e. not provided by an add-on package) time series structure is that of `ts`, for regular (univariate or multivariate) time series. `ts` objects are a simple vector (or matrix) with a class and time attributes:
```{r}
attributes(co2)
```
where `tsp` denotes start, end, and frequency. If frequency is 12 or 4, R uses some heuristics to interpret these data as monthly and quarterly, respectively: 

```{r}
(m.ts = ts(1:3, start = mo.ny[1], end = mo.ny[3], frequency = 12))
(q.ts = ts(1:5, start = qua.ny[1], end = qua.ny[5], frequency = 4))
```

### zoo, xts

Packages `zoo` and `xts` provide ways to deal with data that have `Date` or `POSIXct` time stemps, and that are possibly irregular:
```{r}
library(xts)
(z = zoo(rnorm(3), mo.d))
(x = xts(rnorm(3), mo.d))
```
They have functions to find out whether time series are regular:
```{r}
is.regular(x)
is.regular(x, strict = TRUE)
```