# Time series: identifying frequencies 

Here are two links that try to explain the fourier transform:

* https://betterexplained.com/articles/an-interactive-guide-to-the-fourier-transform/
* http://toxicdump.org/stuff/FourierToy.swf


### The Fourier transform

The [Fourier transform](https://en.wikipedia.org/wiki/Fourier_series)
decomposes any signal into the weighted sum of a number of sin and cos
functions with different frequency:

$$ s_N(t) = a_0/2 + \sum_{n=1}^{N} \big( a_n \cos(\frac{2\pi n t}{P}) + b_n sin(\frac{2\pi n t}{P}) \big) $$

with $P$ the period (frequency $1/P$), and $N$ the length of the series.
The Fourier transform, or spectrum, or periodogram, depicts amplitude
against frequency.

### artificial data:

pure sinus data, unsmoothed (raw) spectrum:
```{r fig=TRUE}
n = 1000
x = pi * 1:n
s1 = sin(x / 100)
plot(s1, type = 'l')
plot(spectrum(s1))
```

smoothed spectrum:

```{r}
plot(spectrum(s1, spans = c(3,5)))
```

higher frequency sinus:

```{r}
s2 = sin(x / 10)
plot(s2, type = 'l')
plot(spectrum(s2))
```

smoothed curve:

```{r}
plot(spectrum(s2, spans = c(3,5)))
```

superposition of both:

```{r}
plot(s1+s2, type = 'l')
plot(spectrum(s1+s2, spans = c(3,5)))
```

adding noise to them

```
ns = s1 + s2 + rnorm(n)
plot(ns, type = 'b')
plot(spectrum(ns))
```

smoothed spectrum:

```{r}
plot(spectrum(s1+s2+rnorm(n), spans = c(3,5)))
```

### stl components of co2:

```{r fig=TRUE}
plot(co2)
s = stl(co2, 12)
plot(s)
s$time.series[1:10,]
```

fetching the seasonal component:

```{r fig=TRUE}
seasonal = s$time.series[,1]
plot(seasonal)
plot(spectrum(seasonal))
plot(spectrum(seasonal, spans = c(3,5)))
```

fetching the trend component:

```{r fig=TRUE}
trend = s$time.series[,2]
plot(trend)
plot(spectrum(trend))
```

fetching the remainder:

```{r fig=TRUE}
remainder = s$time.series[,3]
plot(remainder)
plot(spectrum(remainder))
```
