# correlations, cross correlations, spatial correlation"

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Irish wind dataset

Consider the classic Irish wind dataset
([Haslett and Raftery, 1989](https://www.jstor.org/stable/2347679?seq=1#metadata_info_tab_contents)):
```{r}
library(gstat)
data(wind)
ls()
head(wind)
head(wind.loc)
```

Let's take a single vector out of the `wind` dataset  , and plot it
```{r}
rpt = wind$RPT
plot(rpt, type = 'l')
```

We now see that the x-axis is not meaningfully referenced (index: 1,2,...,n); we can do better by defining a `ts` object:
```{r}
rpt = ts(wind$RPT, start = 1961, frequency = 365.25)
plot(rpt)
rpt = ts(wind$RPT[1:365], start = 1961, frequency = 365.25)
plot(rpt)
```


```{r}
rpt = ts(as.matrix(wind), start = 1961, frequency = 365.25)
plot(rpt[,4:13])
plot(rpt[,4])
```

What is the correlation between two stations?
```{r}
library(maptools) # char2dms
wind.loc[1:3,]
wind.loc$y = as.numeric(char2dms(as.character(wind.loc[["Latitude"]])))
wind.loc$x = as.numeric(char2dms(as.character(wind.loc[["Longitude"]])))
library(maps) # map()
library(mapdata) # worldHires
map("worldHires", xlim = c(-11,-5.4), ylim = c(51,55.5)) # creates plot
library(sf)
wind.loc = st_as_sf(wind.loc, coords = c("x", "y"), crs = 
											"+proj=longlat +datum=WGS84")
plot(wind.loc, add=TRUE, pch=16, col = 'grey')
text(st_coordinates(wind.loc), pos=1, label=wind.loc$Station, cex = .75, col = 'blue')
```

Correlations between stations, matching by time:
```{r}
print(cor(wind[,-(1:3)]), digits = 2)
```

```{r}
names(wind)
acf(wind[,c("MUL", "DUB")])
acf(wind[,c("MUL", "DUB", "BIR", "KIL")])
acf(wind[,-(1:3)])
```

Which values do the stations have, at a particular time, or averaged over time?
```{r}
map("worldHires", xlim = c(-11,-5.4), ylim = c(51,55.5))
plot(wind.loc, add=TRUE, pch=16)
# merge wind and wind.loc:
df = setNames(data.frame(wind = t(wind[2,-(1:3)])), "wind")
df$Code = row.names(df)
out = merge(wind.loc, df)
text(st_coordinates(out), pos=1, label=as.character(out$wind), cex = .75, col = 'blue')
```

```{r}
m = map("worldHires", xlim = c(-11,-5.4), ylim = c(51,55.5), 
				fill = TRUE, plot = FALSE)
plot(out["wind"], reset = FALSE, pch = 16, cex = 3, key.pos = 2,
  xlim = c(-11,-5.4), ylim = c(51,55.5))
plot(st_as_sfc(m), add = TRUE)
```
