```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Geostatistical prediction

Linear predictors:
$$\hat{Z}(s_0) = \sum_{i=1}^n \lambda_i Z(s_i)$$

Precition weights are such that the prediction error $\hat{Z}(s_0)-Z(s_0)$ has expectation zero and minimal variance; this involves inverting the covariance matrix of $Z=\{Z(s_1),...,Z(s_n)\}$.

## Demo case:

### Model variogram:
```{r}
library(sf)
demo(meuse_sf, ask = FALSE)
library(gstat)
v = variogram(log(zinc)~1, meuse_sf)
v.fit = fit.variogram(v, vgm(1, "Sph", 900, 1))
plot(v, v.fit)

demo(meuse, package = "sp")
library(stars)
meuse_grid = st_as_stars(meuse.grid)
```

### Use in prediction:
```{r}
k = krige(log(zinc)~1, meuse_sf, meuse_grid, v.fit)
plot(k["var1.pred"], main = "log(zinc) predictions", breaks = "equal", reset = FALSE)
plot(meuse_sf, add = TRUE, col = 'green', pch = 16)
```

### How to create a prediction grid like `meuse_grid` from scratch?

E.g. from a polygon, e.g. a shapefile read by `rgdal::readOGR`. Here, we'll start with `meuse.area`
```{r}
data(meuse.area, package = "sp")
head(meuse.area) # the points of a polygon
plot(meuse.area, type = 'l', asp = 1)
x = st_as_stars(st_sfc(st_polygon(list(meuse.area))))
plot(x)
x[x==0]=NA
plot(x)
```

## Next:

read all vignettes of the gstat package except for vignette ``prs'':
```{r eval = FALSE}
vignette(package = "gstat")
vignette("gstat")
```
Try to run the code yourself with

```{r eval=FALSE}
edit(vignette("gstat"))
```

and copy-and-paste.

## Exercises

E1. Import one of the pm10 trajectories from http://pebesma.staff.ifgi.de/pm10/ using `read.csv`

```{r}
pm = read.csv("http://pebesma.staff.ifgi.de/pm10/3_1a.txt")
```

E2. Load package `sf` and use `st_as_sf` to convert this into an `sf` object, using argument `coords`; pay attention to the order of coordinates, and set `crs=4326` to specify longlat

```{r}
library(sf)
pm.sf = st_as_sf(pm, coords = c("Long", "Lat"), crs = 4326)
```

E3. Convert to EPSG 32632 (UTM zone 32N), using `st_transform`

```{r}
pm.sf_utm = st_transform(pm.sf, 32632)
```

E4. Load package `gstat` and compute a variogram of the pm10 variable, with maximum distance 300 m

```{r}
library(gstat)
v = variogram(pm10 ~ 1, pm.sf_utm, cutoff = 300)
plot(v)
```

E5. Fit a variogram model to this, using `vgm(1, "Exp", 100, 1))` as the initial fit; save the fitted model

```{r}
v.fit = fit.variogram(v, vgm(1, "Exp", 100, 1))
v.fit
plot(v, v.fit)
```

E6. Load package `stars` and compute a regular grid over the area, by first converting the `st_bbox` of your data to an `sfc` polygon using `st_as_sfc`, and then `st_as_stars` of that polygon.
```{r}
plot(pm.sf_utm["pm10"])
pol = st_as_sfc(st_bbox(pm.sf_utm))
plot(pol)
plot(pm.sf_utm["pm10"], add = TRUE)
grd = st_as_stars(pol)
plot(grd)
grd
```

E7. Carry out a kriging using formula `pm10~1` (constant mean model), the (transformed) data, the grid, and the fitted model

```{r}
kr = krige(pm10~1, pm.sf_utm, grd, v.fit)
```

E8. Plot the resulting kriging predictions, and try to plot the observation (pm10) points on top of it
```{r}
plot(kr[1], breaks = "equal", reset = FALSE)
# plot(st_geometry(pm.sf_utm), add = TRUE, col = 'red', pch = 3, cex = 0.5)
```

E9. Plot the resulting kriging variances

```{r}
plot(kr[2], breaks = "equal")
```
