# Spatial data: introduction

Similar to time series data, spatial data come in at least three fundamentally different forms:

* point patterns: here, the locations carry the essential information: accidents, storms, earthquakes
* geostatistical data (fields): the measured attribute is of interest: air quality, temperature, anything that varies continuously over space
* lattice data: measured values reflect aggregates over large areas (population density per state, number of disease cases per hospital ward, average surface temperature over a 250 x 250 m grid cell), average yearly temperature increase by continent

We also ask different questions to these data:

* point patterns: are there differences in the density/intensity, or is it completely spatial random? Are there second-order effects such as repulsion or attraction between points?
* geostatistical data: can we interpolate the variable at locations where it was not measured? can we compute averages, or (temporal) trends? Can we simulate from the data generating process?
* lattice data: are neighbouring regions spatially correlated? (How) can we run regression models on them, taking into account the spatial correlation?


## geostatistical data.

Suppose we have spatial data, distributed over a 10 x 10 grid, with spacing 100 m:
```{r fig=TRUE}
sq = seq(100,1000,100)
grd = expand.grid(x = sq, y = sq)
icov = exp(-as.matrix(dist(grd))/300)
set.seed(13333)
grd$z = t(rnorm(100) %*% chol(icov))
library(sp)
coordinates(grd) = ~x+y
spplot(grd)
```

Suppose we are interested in the correlation of two neigbhouring locations. A difficult way would be to form a matrix of z, and shift it right and up:
```{r fig=TRUE}
m = matrix(grd$z, 10, 10)
pts1 = cbind(as.vector(m[-10,]), as.vector(m[-1,]))
pts2 = rbind(pts1, cbind(as.vector(m[,-10]), as.vector(m[,-1])))
plot(pts2, asp=1, main = "lag-1 (100 m) scatter plot of z")
```

Where clearly we see the lag-1 correlation, which in this case is a spatial correlation! We could go on like this for lag-2, lag-3 etc, but then we would miss diagonal directions:
```{r fig=TRUE}
plot(sort(as.vector(spDists(grd))))
```

We can get a larger set of scatterplots, grouped by spatial distance intervals, when using gstat::hscat (lagged scatterplots):
```{r, fig=TRUE}
library(gstat)
hscat(z~1, grd, breaks = c(0,100,150,200,300,400,500))
```

In geostatistics, rather than looking at covariances and correlations, for which the building block is
$$(z(s) - m_1)(z(s+h) - m_2)$$
we rather look at semivariances, for which the building block is
$$((z(s) - z(s+h))^2$$
In particular, the semivariance for a distance (lag) interval $h$ is defined as
$$\gamma(h) = \frac{1}{2}E((Z(s)-Z(s+h))^2)$$
and is estimated from sample data $Z(s_i)$, $i=1,...,n$ by
$$\hat{\gamma}(h) = \frac{1}{2N_h}\sum_{i=1}^{N_h}(Z(s_i)-Z(s_i+h))^2$$
where $N_h$ is the number of point pairs $\{Z(s_i),Z(s_i+h)}$ separated by distance (interval) $h$.

Doing this:
```{r}
v = variogram(z~1, grd, cutoff = 800)
plot(v, plot.numbers = TRUE)
```

In the plot (the sample semivariogram or sample variogram) we see $\hat{\gamma}(h)$ on the $y$-axis, $h$ on the $x$-axis. Unlike correlation which goes down, semivariance goes up as an indication of less autocorrelation, and settles on a plateau, similar to the variance of $Z$. 

## The variogram model

For interpolation, we need semivariance values at _every_ value for $h$. This requires fitting a model to the sample variogram. Not every model works well, but a number are known to work well. We'll try the exponential here:
```{r plot=TRUE}
v.fit = fit.variogram(v, vgm("Exp"))
v.fit
plot(v, v.fit)
```

Once we "have" this variogram model, we can do pretty much everything: interpolation, simulation, averaging, ...
