```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Geostatistical prediction

Linear predictors:
$$\hat{Z}(s_0) = \sum_{i=1}^n \lambda_i Z(s_i)$$

Precition weights are such that the prediction error $\hat{Z}(s_0)-Z(s_0)$ has expectation zero and minimal variance; this involves inverting the covariance matrix of $Z=\{Z(s_1),...,Z(s_n)\}$.

## Demo case:

### Model variogram:
```{r}
library(sf)
demo(meuse_sf, ask = FALSE)
library(gstat)
v = variogram(log(zinc)~1, meuse_sf)
v.fit = fit.variogram(v, vgm(1, "Sph", 900, 1))
plot(v, v.fit)

demo(meuse, package = "sp")
library(stars)
meuse_grid = st_as_stars(meuse.grid)
```

### Use in prediction:
```{r}
k = krige(log(zinc)~1, meuse_sf, meuse_grid, v.fit)
plot(k["var1.pred"], main = "log(zinc) predictions", breaks = "equal", reset = FALSE)
plot(meuse_sf, add = TRUE, col = 'green', pch = 16)
```

### How to create a prediction grid like `meuse_grid` from scratch?

E.g. from a polygon, e.g. a shapefile read by `rgdal::readOGR`. Here, we'll start with `meuse.area`
```{r}
data(meuse.area, package = "sp")
head(meuse.area) # the points of a polygon
plot(meuse.area, type = 'l', asp = 1)
x = st_as_stars(st_sfc(st_polygon(list(meuse.area))))
plot(x)
x[x==0]=NA
plot(x)
```

## Next:

read all vignettes of the gstat package except for vignette ``prs'':
```{r eval = FALSE}
vignette(package = "gstat")
vignette("gstat")
```
Try to run the code yourself with

```{r eval=FALSE}
edit(vignette("gstat"))
```

and copy-and-paste.

## Exercises

E1. Import one of the pm10 trajectories from http://pebesma.staff.ifgi.de/pm10/ using `read.csv`

E2. Load package `sf` and use `st_as_sf` to convert this into an `sf` object, using argument `coords`; pay attention to the order of coordinates, and set `crs=4326` to specify longlat

E3. Convert to EPSG 32633 (UTM zone 33N), using `st_transform`

E4. Load package `gstat` and compute a variogram of the pm10 variable, with maximum distance 300 m

E5. Fit a variogram model to this, using `vgm(1, "Exp", 100, 1))` as the initial fit; save the fitted model

E6. Load package `stars` and compute a regular grid over the area, by first converting the `st_bbox` of your data to an `sfc` polygon using `st_as_sfc`, and then `st_as_stars` of that polygon.

E7. Carry out a kriging using formula `pm10~1` (constant mean model), the (transformed) data, the grid, and the fitted model

E8. Plot the resulting kriging predictions, and try to plot the observation (pm10) points on top of it

E9. Plot the resulting kriging variances
