# Spatial data: interpolation, simulation, averaging

### Variogram, covariogram, auto-correlogram

```{r, fig=TRUE}
library(sf)
demo(meuse_sf, ask = FALSE) # load meuse dataset
library(gstat)
hist(meuse$zinc)
hist(log(meuse$zinc))
plot(variogram(log(zinc)~1, meuse_sf))
```

How can we relate the variogram to the autocorrelogram?

1. There is no scaling to $[-1,1]$ (this scaling converts covariances to correlation)
2. Variogram values are strictly positive (mean squares)
3. Variogram increase with distance (with decreasing correlation)

The variogram and covariogram relate through
$$\gamma(h) = C(0) - C(h)$$
with $C(h)$ the covariogram of the process:
```{r fig=TRUE}
plot(variogram(log(zinc)~1, meuse_sf, covariogram = TRUE))
```

### What to choose as lag distance?
```{r fig=TRUE}
plot(variogram(log(zinc)~1, meuse_sf))
plot(variogram(log(zinc)~1, meuse_sf, width = 50))
plot(variogram(log(zinc)~1, meuse_sf, width = 250))
```

All still somewhat acceptible; things get weirder when
```{r fig=TRUE}
plot(variogram(log(zinc)~1, meuse_sf, width = 20))
plot(variogram(log(zinc)~1, meuse_sf, width = 500))
```

### Variogram model
For further analysis, we need a variogram _model_, a parametric model $\gamma(h)$ that gives semivariance (and hence: covariance, autocorrelation) values for _every_ value of $h$:
```{r}
v = variogram(log(zinc)~1, meuse_sf)
f = fit.variogram(v, vgm(1, "Sph", 300, 1))
plot(v, f)
f
```

The reason for this is that we would like to be able to make a prediction (interpolation) for _any_ location, and this involves _any_ distance(s) to observation values.

Compare: for time series we typically predict (forecast) $y_{t+1\delta}, y_{t+2\delta},...$. Would we want to predict (forecast) values for values _inbetween_, like $y_{t+0.33\delta}$ or $y_{t+7.51\delta}$, then we'd need autocorrelations for these non-multiples of the time interval $\delta$, and we'd need to _interpolate_ (model) the autocorrellogram. 

### Positive definite

Models we fit to a covariogram (or autocorrelation) need to be _positive definite_. This means that for an arbitrary set of points $p_1,...,p_2$ the covariance matrix $C$ of the these points has the following property
$$a' C a > 0$$
for non-zero but otherwise arbitrarily chosen weights coefficients $a$. This guarantees, amongst others, that prediction variances are non-negative.

For variograms, signs flip, and we require _negative definite_ functions, to the same effect.

It is easy to go wrong here, for instance the break-line model:
```{r}
f0 = fit.variogram(v, vgm(1, "Lin", 700, 1))
plot(v, f0)
```

does have this property for points in 1, but not in 2 or more dimensions!

### Coordinates in lat/long

* coordinates on the sphere (geographic, geodetic coordinates) are expressed as angles (longitude, latitude), and not as distances
* hence, to compute distances between points ($h$), we cannot use Euclides' formula $\sqrt{(x_1-x_2)^2+(y_1-y_2)^2}$
* R can deal with this automatically, but you need to tell it!

```{r}
meuse_ll = st_transform(meuse_sf, "+proj=longlat +datum=WGS84")
plot(variogram(log(zinc)~1, meuse_ll), main = "distance in km!!")
st_crs(meuse_ll) = NA # remove CRS
plot(variogram(log(zinc)~1, meuse_ll), main = "distance units ?????")
```

More in detail:
```{r}
(pts = rbind(c(0,0), c(1,0), c(0,1)))
as.matrix(dist(pts))
df = data.frame(pts)
names(df) = c("x", "y")
st_distance(st_as_sf(df, coords = c("x", "y"), crs = NA))
st_distance(st_as_sf(df, coords = c("x", "y"), crs = "+proj=longlat"))
(pts_here = rbind(c(7,52), c(8,52), c(7,53)))
df_here = data.frame(pts_here)
names(df_here) = c("x", "y")
st_distance(st_as_sf(df_here, coords = c("x", "y"), crs = NA))
st_distance(st_as_sf(df_here, coords = c("x", "y"), crs = "+proj=longlat"))
```

Note that:

* Euclidian distance calculations are insensitive for swapping x and y
* Geodetical distance calculations are **NOT** insenstive to axis swapping
* it is of crucial importance to _not_ mix up longitude and latitude
* R adopts this order: 1. longitude, 2. latitude (think: x, y)

```{r}
st_distance(st_as_sf(df_here, coords = c("x", "y"), crs = NA))
st_distance(st_as_sf(df_here, coords = c("y", "x"), crs = NA))
st_distance(st_as_sf(df_here, coords = c("x", "y"), crs = "+proj=longlat"))
st_distance(st_as_sf(df_here, coords = c("y", "x"), crs = "+proj=longlat"))
```
