# Spatial Point Patterns

Recommended reading: _Statistical Analysis of Spatial Point Processes_, chapter 7 in _Applied Spatial Data Analysis with R, second edition_ ; Springer (2013).

Very nice slides from a point process expert are found [here](http://www.icms.org.uk/downloads/spat_eco/illian.pdf); Illian is also first author of [this book](http://eu.wiley.com/WileyCDA/WileyTitle/productCd-0470014911.html)

Spatial Point Pattern analysis is concerned with describing the
distribution of points in space; usually a 2-D Euclidian (metric)
space is assumed.

Over a give window $W$, we observe a number of $n$ events (points)  $S_i$, $i=1,...n$.

We distinguish:

* a point pattern: the observed locations of events in their window
* a point process: the stochastic process that generates point pattern as its outcome

We can describe:

* first order properties: concerning the density of the events (is the density of the point pattern constant over the window?)
* second order properties: concerning the interaction between events (do events attract each other, repulse each other, or act indifferently with respect to each other?)

The baseline process is _completely spatial random_ (CRS): it means that the density of the process is constant, and points occur completely independently, meaning there is no interaction.

Testing for constant density can be done with the quadrat counts test:
* our point pattern has $N$ events
* cut the area in n boxes of equal size
* count the number of events per box, $n_i$
* the sum of $(n_i - N/n)^2$ follows a $\chi^2$ distribution with $n-1$ degrees of freedom under the null hypothesis of CSR
Other tests involve e.g. distances between points.

Caveats:
* how to determine $n$?
* can we get equally sized boxes in real cases?
* as usual: not significant does not mean that the process _is_ CSR, it can also indicate lack of power (small sample)

Alternatives to CSR: 
* attracting or repulsing processes, constant density (e.g. Strauss)
* no interaction process, varying density (log Gaussian Cox process)

Software:
* spatial: [spatstat](https://spatstat.github.io/) (JSS papers :[1](http://www.jstatsoft.org/v12/i06/), [2](http://www.jstatsoft.org/v55/i11/))
* spatio-temporal: CRAN: [stpp](https://cran.r-project.org/package=stpp) [JSS paper](https://www.jstatsoft.org/article/view/v053i02); [lgcp](https://cran.r-project.org/web/packages/lgcp/) JSS papers: [1](http://www.jstatsoft.org/v52/i04/), [2](http://www.jstatsoft.org/v63/i07/)

Typical spatio-temporal questions involve whether the density (if varying) varies over time in a way that is independent from the variation over space. In case of an epidemics, movement of the epidemic would imply interaction (FMD examples in the stpp package/paper).

Functions:
* K-function (Ripley's K): the average number of other points found within the distance $r$ from the typical point.
* L-function: similar to K, but CSR is a line rather than a parabolic curve
* g-function: (pair-correlation function); recommended by Illian et al., similar to L and K but $g(r) = 1$  for CSR cases

Inference:
* inference often follows a simulation approach: point patterns are simulated given a particular process, and the correspondence of the (set of) simulations to the observed processed is considered.

```{r}
library(spatstat)
data(japanesepines)
summary(japanesepines)

library(maptools)
spjpines <- as(japanesepines, "SpatialPoints")
summary(spjpines)

spjpines1 <- elide(spjpines, scale=TRUE, unitsq=TRUE)
summary(spjpines1)

pppjap <- as(spjpines1, "ppp")
summary(pppjap)

data(redwoodfull)
spred <- as(redwoodfull, "SpatialPoints")
data(cells)
spcells <- as(cells, "SpatialPoints")
dpp<-data.frame(rbind(coordinates(spjpines1), coordinates(spred), 
   coordinates(spcells)))
njap<-nrow(coordinates(spjpines1))
nred<-nrow(coordinates(spred))
ncells<-nrow(coordinates(spcells))
dpp<-cbind(dpp,c(rep("JAPANESE",njap), rep("REDWOOD", nred), rep("CELLS", ncells))) 
names(dpp)<-c("x", "y", "DATASET")


library(lattice)
print(xyplot(y~x|DATASET, data=dpp, pch=19, aspect=1))

setwd("./asdar")

library(rgdal)
spasthma <- readOGR(".", "spasthma")
spbdry <- readOGR(".", "spbdry")
spsrc <- readOGR(".", "spsrc")
sproads <- readOGR(".", "sproads")

```

```{r}
setwd("../astd")

plot(spbdry, axes=TRUE, lwd=0.5)
plot(sproads, add=TRUE, lwd=2, col="darkslategrey")
c_c <- (spasthma$Asthma == "case") + 1
plot(spasthma[c_c == 1,], add=TRUE, pch=4, cex=0.6, col="mediumaquamarine")
plot(spasthma[c_c == 2,], add=TRUE, pch=17, cex=0.75, col="goldenrod2")
plot(spsrc, pch=22, add=TRUE, cex=1.2, bg="brown4")
legend("bottomright", legend=c("controls", "cases", "pollution sources"), pch=c(4, 17, 22), pt.cex=c(0.6, 0.75, 1.2), pt.bg=c(NA, NA, "brown4"), col=c("mediumaquamarine", "goldenrod2", "black"), bty="n") 


## envjap <- envelope(as(spjpines1, "ppp"), fun=Gest, r=r, nrank=2, nsim=99)
## envred <- envelope(as(spred, "ppp"), fun=Gest, r=r, nrank=2, nsim=99)
## envcells <- envelope(as(spcells, "ppp"), fun=Gest, r=r, nrank=2, nsim=99)
## Gresults <- rbind(envjap, envred, envcells) 
## Gresults <- cbind(Gresults, 
##    y=rep(c("JAPANESE", "REDWOOD", "CELLS"), each=length(r)))
## # CHANGED DATASET TO y RSB
## #	save(Gresults, envjap, envred, envcells, file="sppaGestEnv.RData")


###################################################
### code chunk number 20: sppa.Rnw:559-560
###################################################
#cat("%", difftime(Sys.time(), tic, units="secs"), "seconds\n\n")


###################################################
### code chunk number 21: sppa.Rnw:562-563
###################################################
load("Gresults.RData")

print(xyplot(obs~theo|y , data=Gresults, type="l", 
xlab = "theoretical", ylab = "observed", # EJP
panel=function(x, y, subscripts) {
   lpolygon(c(x, rev(x)), 
   c(Gresults$lo[subscripts], rev(Gresults$hi[subscripts])),
   border="gray", col="gray"
)
llines(x, y, col="black", lwd=2)
}))


```
