---
title: "geostatistics: 2"
output: html_document
---
# Geostatistics, continued

## What is the actual model?
In the case of
```{r}
library("sp")
data(meuse)
coordinates(meuse) = ~x+y
data(meuse.grid)
coordinates(meuse.grid) = ~x+y
gridded(meuse.grid) = TRUE
library("gstat")
model = vgm(.6, "Sph", 900, .05)
k = krige(log(zinc)~1, meuse, meuse.grid, model)
spplot(k["var1.pred"])
```

the actual model used for the $\log(\mbox{zinc})$ data, called $Z(s)$ is:
$$Z(s) = m + e(s)$$
with $m$ a (spatially) constant mean, and $e(s)$ a residual for which
$$E(e(s)) = 0$$
and
$$0.5 Var(e(s)-e(s+h)) = \gamma(h)$$
with $\gamma(h)$ the variogram _model_ of $Z(s)$:

```{r}
v = variogram(log(zinc)~1, meuse)
plot(v, model)
```

Here, `v` is the _sample variogram_, to which the variogram model is usually fitted:

```{r}
fit.variogram(v, model)
```

This model gives us values of the semivariance, and by that the covariance, for _every distance_, thus enabling interpolation at _any location_.

We distinguish the first order properties ($m$, mean: spatially constant) from the second order propertiers ($\gamma(h)$ variogram: spatially stationary, meaning independent of location $s$, only depending on separation distance)

We can choose more elaborate models for the trend _function_ by specifying predictors in the formula, e.g. 

```{r}
k2 =krige(log(zinc)~dist, meuse, meuse.grid, model)
```

uses the model
$$Z(s) = \beta_0 + \beta_1 f(s) + e(s)$$
with $f(s)$ the distance to river at location $s$. This variable needs to be present at all locations in both `meuse` and `meuse.grid`, in order to carry out prediction.

## What can we do to improve?

Note that we already improved by working on log-concentrations rather than raw concentrations.

Some other options:

1. transform the distribution: by log, or by some other form (square root? cube root? power transform? inverse Gaussian? Box-Cox?)
1. carry out local kriging instead of global kriging
1. make the variogram direction-dependend (assume anisotropy, instead of isotropy)
1. stratify the region, and carry out variogram modelling and interpolation for each stratum (sub-region)

Example of local kriging:

```{r}
k$var1.pr.n5 = krige(log(zinc)~1, meuse, meuse.grid, model, nmax = 5)$var1.pred
spplot(k[c("var1.pred", "var1.pr.n5")], labels = c("global", "local, n=5"))
```


## How can we evaluate, and compare models?

Evaluation usually includes _visual_ control, i.e. is the predicted surface realistic, and fit for the purpose?

Other options: cross validation

```{r}
cv = krige.cv(log(zinc)~1, meuse, model, nmax = 5)
```

mean errors:

```{r}
mean(cv$residual)
```

error variability, e.g. RMSE:
```{r}
sqrt(mean(cv$residual^2))
```


## What do prediction errors mean?
Consider at an arbitrary, single location

```{r}
pt = SpatialPoints(cbind(179660, 331860))
image(k[1])
points(pt)
```

the prediction and prediction error:

```{r}
pr = as.data.frame(k[pt,])
pr
```

Assuming that the prediction error is normally distributed, we can form a 95% prediction interval for $Z(s)$ at this point by

```{r}
qnorm(c(0.025, 0.975), mean = pr$var1.pred, sd = sqrt(pr$var1.var))
```

which, transformed to the original scale yields a 95% prediction interval for the _median_ value:

```{r}
exp(qnorm(c(0.025, 0.975), mean = pr$var1.pred, sd = sqrt(pr$var1.var)))
```

We can form maps of both sides of the interval:

```{r}
k$lower = exp(qnorm(0.025, mean = k$var1.pred, sd = sqrt(k$var1.var)))
k$upper = exp(qnorm(0.975, mean = k$var1.pred, sd = sqrt(k$var1.var)))
spplot(k[c("lower", "upper")])
```

and we could classify intervals according to their position with respect to a threshold value, e.g. 500:

```{r}
f = ifelse(k$upper < 500, 1, ifelse(k$lower > 500, 3, 2))
k$f = factor(f, labels = c("lower", "not distinguishable", "higher"))
spplot(k["f"], main = "zinc concentration prediction relative to 500 ppm")
```

## Multivariate geostatistics

Just as the _cross correlelation function_ gives the (lagged, cross) correlation between two time series, the _cross variogram_ gives the spatial cross correlation between two spatial variables. In gstat, we build a `gstat` object holding one _or more_ variables, by incrementally adding one variable:

```{r}
g = gstat(NULL, "log-zn", log(zinc)~1, meuse)
g = gstat(g, "log-pb", log(lead)~1, meuse)
g = gstat(g, "log-cd", log(cadmium)~1, meuse)
v = variogram(g)
plot(v)
```

Fitting a model to these is somewhat constrained, and needs to be jointly; the following fits a _linear model of coregionalisation_ (LMC):

```{r}
v.fit = fit.lmc(v, g, vgm(1, "Sph", 900, 1))
plot(v, v.fit)
```

Prediction under such models, called _cokriging_, is done _jointly_, i.e. the three variables $Z_1(s)$, $Z_2(s)$ and $Z_3(s)$ are predicted in a single step, that also results in the prediction covariance matrix with the prediction variances and covariance for a particular location.

## Conditional simulation
Kriging and cokriging provide the _marginal_ distribution of $Z(s_0)$ (with scalar or vector $Z$), but not the joint distribution of $Z(\cdot)$ at different locations. Meaning, we cannot answer the question, for two locations $p$ and $q$, what the probability is that _both_ $Z(p)$ and $Z(q)$ exceed a threshold. The multivariate distribution needed is high-dimensional if we consider many prediction locations, and so we usually _sample_ from this distribution, and call this _conditional simulation_

```{r}
s = krige(log(zinc)~1, meuse, meuse.grid, model, nsim = 10, nmax = 15)
spplot(s)
```

Unlike predictions which are always smooth, the variogram of simulations resembles that of the data:

```{r}
v1 = variogram(var1.pred~1, k)
v2 = variogram(sim1 ~ 1, s)
plot(gamma~dist, v2, xlim = c(0, 1500), ylim = c(0, 0.7))
points(gamma~dist, v1, col = 'red')
lines(variogramLine(model, 1500, col = 'blue'))
```

## Geostatistics for temporally varying data

Spatial, temporally varying data can be seen, most simply as coming from the model
$$Z(s,t) = m + e(s,t)$$
with $s$ and $t$ the spatial location and temporal index (instance) of $Z$.

Suppose that we have only temporal replications for a small number of time steps, $t_i, i = 1,...,k$. We could then treat $Z(s,t)$ as $Z(s,t_i) = Z_{i}(s)$. If we are not interested in predictions for other time steps, we can proceed as if the time steps constitute different variables, and treat it as a multivariable problem. Run, for instance, `demo(pcb)`.

A simple alternative would be to treat $(s,t)$ as a three- or (four-) dimensional space, model variograms and interpolate in this space. More
alternatives will follow later.

## Exercises

1. run demo `pcb` in package `gstat`, and try to understand what is going on.
1. find a data sets _not used in one of the R packages_ that allows for geostatistical interpolation.
1. interpolate the data using some kriging variety.
1. give results (variogram, map) and some model diagnostics.
