reading data from the airbase data sets at
http://www.eea.europa.eu/data-and-maps/data/airbase-the-european-air-quality-database-7
data by country -> download file -> unzip file -> unzip AirBase_XX_v7_rawdata.zip with XX
the iso code
1: daily data
```{r}
library(xts)
# get all files:
#l = list.files(pattern = "DE")
l = list.files(pattern = "DK")
# select those with daily data:
days = l[substr(l, 18, 20) == "day"]
# code 5:
pm10files = days[substr(days, 8, 12) == "00005"]
```

A function that reads in a single month:
```{r}
read_month <- function (filename) {
	f0 = read.table(filename)
	f0$Date = as.Date(f0[[1]])
	# select the right columns:
	f = f0[,seq(2, 62, by = 2)]
	# stack the rows in colums
	f = reshape(f, varying = list(1:31), direction = "long")
	# now, id time day, and id is the month index:
	f$Date = f0$Date[f$id] + f$time - 1
	# get the quality flags:
	qual = reshape(f0[,seq(3, 63, by = 2)], varying = list(1:31), direction = "long")
	# deselect everything with quality flag == 0: this also excludes the 
	# non-existing days (e.g. day number 31 for even months)
	f = f[qual$V3 != 0,]
	# create an xts file, return it:
	xts(f$V2, f$Date) 
}
```
Apply this function to every file, return a list:
```{r}
lst = lapply(pm10files, function(x) read_month(x))
```
merge all elements of this list:
```{r}
daily.pm10 = do.call(merge, lst)
# add station ID as variable names to the new object:
dimnames(daily.pm10) = list(NULL, substr(pm10files, 1, 7))
head(daily.pm10)
```

## 2: hourly data

```{r}
days = l[substr(l, 18, 21) == "hour"]
# code 5:
pm10files = days[substr(days, 8, 12) == "00005"]
```

read a the hourly data for each single station:
```{r}
Sys.setenv(TZ = "UTC") # avoid DST problems...
read_day <- function(filename) {
	f0 = read.table(filename)
	f0$time = as.POSIXct(f0[[1]])
	# select the right columns:
	f = f0[,seq(2, 48, by = 2)]
	# stack the rows in colums
	f = reshape(f, varying = list(1:24), direction = "long")
	# now, id time day, and id is the month index:
	f$time = f0$time[f$id] + (f$time - 1) * 3600
	# get the quality flags:
	qual = reshape(f0[,seq(3, 49, by = 2)], varying = list(1:24), direction = "long")
	# deselect everything with quality flag == 0: this also excludes the 
	# non-existing days (e.g. day number 31 for even months)
	f = f[qual$V3 != 0,]
	# create an xts file, return it:
	xts(f$V2, f$time) 
}
```
apply this function to all files:
```{r}
lst = lapply(pm10files, function(x) read_day(x))
hourly.pm10 = do.call(merge, lst)
```
there might be stations completely empty (DE):
```{r}
empty = which(sapply(lst, function(x) length(x)==0))
if (length(empty) > 0)
    pm10files = pm10files[-empty]
```
add column names:
```{r}
dimnames(hourly.pm10) = list(NULL, substr(pm10files, 1, 7))
head(hourly.pm10)
```

### NOTE the following:
1. no for loops, but lapply, which returns station datasets in a list
2. do.call is a single call to merge, passing all files as arguments
3. without using dimnames, the resulting object prints VERY messy, as default dimnames are created from the call
4. alternatives to reshape: 
  - `stack`
  - `as.vector(as.matrix(f[,seq(3, 63, by = 2)]))`    # hour-by-hour
  - `as.vector(t(as.matrix(f[,seq(3, 63, by = 2)])))` # in time order
