# Data for final assignment

Data can be downloaded from https://uni-muenster.sciebo.de/s/yvFiuWGqG7riliA/download

This is a zip file (of 25 Gb: takes quite a while); unpacked it is not much larger, as
it contains .tif files.

A script that summarizes the data to 250 m pixels and yearly aggregated (median) NDVI values is:

```{r}
library(gdalcubes)
library(magrittr)
gdalcubes_options(threads=8)

IMAGE_DIR = "L8_cropped" # please change

col = create_image_collection(list.files(IMAGE_DIR, recursive = TRUE, pattern=".tif", full.names  = TRUE), "L8_SR")

# only use "clear" pixels
L8.clear_mask = image_mask("PIXEL_QA", values=c(322, 386, 834, 898, 1346, 324, 388, 836, 900, 1348), invert = TRUE)

# example yearly data cube at 250m spatial resolution
v = cube_view(srs="EPSG:3857", extent=col, dx=250, dy=250, dt="P1Y", resampling = "average", aggregation = "median")

# calculate NDVI and export as GeoTIFF files at subfolder "L8cube"
raster_cube(col, v, L8.clear_mask) %>%
  select_bands(c("B04", "B05")) %>%
  apply_pixel("(B05-B04)/(B05+B04)") %>%
  write_tif("../L8cube",prefix = "NDVI_")
```

We can read this with package stars, and visualise as:

```{r}
# remotes::install_github("r-spatial/stars")
library(stars)
subdir = "L8cube"
f = paste0(subdir, "/", list.files(subdir))
(st = read_stars(f))
plot(merge(st))
plot(merge(st), breaks = "equal")
```

To see the raw (non-aggregated) data, with all the cloud gaps, create a
composite for a sub-region by
```{r}
gdalcubes_options(threads=8)

IMAGE_DIR = "L8_cropped" # please change

col = create_image_collection(list.files(IMAGE_DIR, recursive = TRUE, pattern=".tif", full.names  = TRUE), "L8_SR")

# only use "clear" pixels
L8.clear_mask = image_mask("PIXEL_QA", values=c(322, 386, 834, 898, 1346, 324, 388, 836, 900, 1348), invert = TRUE)

# example 16 day data cube of a smaller area at 60m spatial resolution
v = cube_view(srs="EPSG:3857", extent=list(left = -7347259, right = -7314864, top = -995476.1, bottom = -1025490, t0 ="2015-01-01", t1 = "2018-12-31"), dx=60, dy=60, dt = "P16D", resampling = "average", aggregation = "median")

# calculate NDVI and export as GeoTIFF files at subfolder "L8cube_subregion"
raster_cube(col, v, L8.clear_mask) %>%
  select_bands(c("B04", "B05")) %>%
  apply_pixel("(B05-B04)/(B05+B04)") %>%
  write_tif("L8cube_subregion",prefix = "NDVI_")
```

And visualise with
```{r}
# remotes::install_github("r-spatial/stars")
library(stars)
subdir = "L8cube_subregion"
f = paste0(subdir, "/", list.files(subdir))
(st = read_stars(f))
plot(merge(st))
```
