---
title: "trajectory + data cube exercise"
author: "Edzer Pebesma"
date: "1/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this exercise, we create a data cube from a set of trajectories

## Data

Consider the storms package in `dplyr`:
```{r}
library(dplyr)
data(storms)
storms
```

The task is analyse the variability of storms, or storm severity, over time. 

Read up on this dataset from its help file, obtained by `?storms`. If you are ambitious, you are also welcome to use the entire hurdat dataset from NOAA.

## Analyse

1. create a time sequence of maps of storm intensity, use (for instance) a spatial grid with 5 degree x 5 degree resolution, and 5-year time periods. Think about whether you want to aggregate storm instances (every storm contributes maximally one data point) or storm fixes, as the 6-hour fixes in the `storms` dataset.

2.  From the time sequence of storms, compute for every grid cell the temporal trend (increase or decrease) of storm intensity, and plot the results.

3. (bonus) repeat 1 and 2, but for each of the levels of storm `status`. 

4. How many dimensions do each of the resulting data cubes have?


```{r}
storms$lat_cl = cut(storms$lat, seq(5, 55, 5), right = FALSE)
storms$long_cl = cut(storms$long, seq(-110, -5, 5), right = FALSE)
storms$yr_cl = cut(storms$year, seq(1975, 2015, 5), right = FALSE)
library(ggplot2)
fn = function(x) length(unique(x))
a = aggregate(storms["name"], storms[c("lat_cl", "long_cl", "yr_cl")], fn)
ggplot(a) + geom_raster(aes(x = long_cl, y = lat_cl, fill = name)) + facet_wrap(~yr_cl)
a = aggregate(storms["year"], storms[c("lat_cl", "long_cl", "yr_cl")], length, drop = FALSE)
ggplot(a) + geom_raster(aes(x = long_cl, y = lat_cl, fill = year)) + facet_wrap(~yr_cl)
a$year [ is.na(a$year) ] = 0
ggplot(a) + geom_raster(aes(x = long_cl, y = lat_cl, fill = year)) + facet_wrap(~yr_cl)
```

```{r}
# try this:
# a = aggregate(storms["year"], storms[c("lat_cl", "long_cl")], print, drop = FALSE)
s = sample(1985:2015, 50, replace = TRUE)
(ts = table(s))
x = as.numeric(names(ts))
y = as.numeric(ts)
plot(y~x)
abline(lm(y~x))
coef(lm(y~x))
coef(lm(y~x))[["x"]]
```

```{r}
get_slope = function(x) {
  ts = table(x)
  X = as.numeric(names(ts))
  y = as.numeric(ts)
  coef(lm(y~X))[["X"]]
}
a = aggregate(storms["year"], storms[c("lat_cl", "long_cl")], get_slope, drop = FALSE)
ggplot(a) + geom_raster(aes(x = long_cl, y = lat_cl, fill = year))
ggplot(subset(a, year > -.7)) + geom_raster(aes(x = long_cl, y = lat_cl, fill = year))
```

Another solution is shown in these slides: https://edzer.github.io/rstudio_conf/2019/#1
with .Rmd found here: https://github.com/edzer/rstudio_conf/blob/master/2019/pres.Rmd
