---
title: "time_series"
author: "Edzer Pebesma"
date: "October 16, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 

```{r}
library(gstat)
data(wind)
ls()
head(wind)
head(wind.loc)
```

Let's take a single vector out of the `wind` dataset, and plot it
```{r}
rpt = wind$RPT
plot(rpt, type = 'l')
```
We now see that the x-axis is not meaningfully referenced (index: 1,2,...,n); we can do better by defining a `ts` object:
```{r}
rpt = ts(wind$RPT, start = 1961, frequency = 365.25)
plot(rpt)
rpt = ts(wind$RPT[1:365], start = 1961, frequency = 365.25)
plot(rpt)
```

## time classes
We can see (from the last plot) that R has no real understanding what the time stands for; we defined it here simply as decimal years, but didn't specify this anywhere. 

R has two native classes for time: `Date` and `POSIXct`:
```{r}
(t0 = Sys.time())
class(t0)
(d0 = Sys.Date())
class(d0)
```
both of which are numeric vectors with a class attribute, where the class attribute points out how the number should be understood, either seconds since 1970-01-01 00:00 GMT
```{r}
as.numeric(Sys.time())
```
or days since that date
```{r}
as.numeric(Sys.Date())
```

## Reading in date or time

We can make time-aware time series objects using packages `zoo` or `xts`:
```{r}
library(xts)
index = paste(1900 + wind$year, wind$month, wind$day, sep = "-")
index[1:3]
index.date = as.Date(index)
index.date[1:3]
as.numeric(index.date[1:3])
rpt = xts(wind$RPT, index.date)
plot(rpt["1961-01::1961-06"])
plot(rpt["1961::1962"])
```

## Autocorrelation

The question whether time series are autocorrelated boils down to whether, for a given time series, the first $n-p$ observations are correlated with the last $n-p$, where $p$ is called the _lag_. We can compute this, by hand, e.g. for the wind data at NRP and lags 1 and 2 by
```{r}
n = length(wind$RPT)
cor(head(wind$RPT, n), tail(wind$RPT, n)) # obviously!
n = length(wind$RPT) - 1
cor(head(wind$RPT, n), tail(wind$RPT, n))
n = length(wind$RPT) - 2
cor(head(wind$RPT, n), tail(wind$RPT, n))
```

When using the R function `acf`, we values that are equal up to four digits,
```{r}
a = acf(wind$RPT, plot = FALSE)
a$acf[1:3]
```
and the possibility to plot the autocorrelations:
```{r}
plot(a)
```
We can plot autocorrelations up to larger lags by
```{r}
plot(acf(wind$RPT, lag.max = 365))
```
Can you explain the negative autocorrelations around lag 180, and positive ones around 360?

## AR(n) processes; generating them

An AR(1) process is defined as $y(t) = \phi_1 y(t-1) + e(t)$ with $phi_1$ the AR(1) coefficient (between 0 and 1), and $e(t)$ independent, normal, random variable.

We can generate an AR(1) by hand by
```{r}
n = 1000
phi = 0.5
y = vector("numeric", n)
y[1] = 0
for (i in 2:n)
	y[i] = phi[1] * y[i-1] + rnorm(1)
plot(y, type = 'l')
```
```{r}
pacf(y)
```
For higher-order AR(n) processes, we have
 $y(t) = \sum_{j=1}^p \phi_j y(t-j) + e(t)$ with $\phi_j = (\phi_1,...,\phi_p)$ the $p$ AR coefficient (between 0 and 1, with additional constraints for $y$ to be stationary), and $e(t)$ independent, normal, random variable.

Although it would be quite easy to modify the code above, it is easier to use `arima.sim` to simulate a particular process, e.g. by
```{r}
y = arima.sim(model = list(ar = c(0.9, -0.1)), n = 1000)
plot(y, type = 'l')
```

Although such series are autocorrelated up to high lags
```{r}
acf(y)
```
it is their _partial_ autocorrelatoin coefficients that reveal their generating model:
```{r}
pacf(y)
```
Partial autocorrelation at lag $p$ ($p > 1$) can be thought of as the autocorrelation between 
the first $n-p$ and the last $n-p$ observations _after correcting for all correlations occuring at shorter lags_.


## ARIMA models

ARIMA models are a very generic class for (linear) time domain analysis. They contain an AR(p) and MA(q) component, and work on $d$ times differenced series:

 $\mbox{Diff}^d (y(t)) = \sum_{j=1}^p \phi_j y(t-j) + \sum_{k=1}^q \alpha_k e(t-k)$ 

```{r}
set.seed(139)
y = arima.sim(model = list(ar = c(0.9, -0.1), order = c(2, 1, 0)), n = 1000)
plot(y, type = 'l')
plot(pacf(y))
plot(pacf(diff(y)))
```

## Exercises thursday:

1. find two time series data sets, of length more than 50
2. Read them in R
3. find a suitable ARIMA model that fits these data
