---
title: "Wind dataset"
author: "Edzer Pebesma"
date: "11/25/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## example(wind)

```{r}
library(gstat)
data(wind)
summary(wind)
wind.loc
library(sp) # char2dms
wind.loc$y = as.numeric(char2dms(as.character(wind.loc[["Latitude"]])))
wind.loc$x = as.numeric(char2dms(as.character(wind.loc[["Longitude"]])))
coordinates(wind.loc) = ~x+y
# fig 1:
if (require(mapdata)) {
    map("worldHires", xlim = c(-11,-5.4), ylim = c(51,55.5))
    plot(wind.loc, add=TRUE, pch=16)
    text(coordinates(wind.loc), pos=1, label=wind.loc$Station)
}
```

```{r}
wind$time = ISOdate(wind$year+1900, wind$month, wind$day)
# time series of e.g. Dublin data:
plot(DUB~time, wind, type= 'l', ylab = "windspeed (knots)", main = "Dublin")
```

```{r}
# fig 2:
#wind = wind[!(wind$month == 2 & wind$day == 29),]
wind$jday = as.numeric(format(wind$time, '%j'))
windsqrt = sqrt(0.5148 * as.matrix(wind[4:15]))
Jday = 1:366
windsqrt = windsqrt - mean(windsqrt)
daymeans = sapply(split(windsqrt, wind$jday), mean)
plot(daymeans ~ Jday)
lines(lowess(daymeans ~ Jday, f = 0.1))
```

```{r}
# subtract the trend:
meanwind = lowess(daymeans ~ Jday, f = 0.1)$y[wind$jday]
velocity = apply(windsqrt, 2, function(x) { x - meanwind })

# match order of columns in wind to Code in wind.loc:
pts = coordinates(wind.loc[match(names(wind[4:15]), wind.loc$Code),])

# fig 3, but not really yet...
dists = spDists(pts, longlat=TRUE)
corv = cor(velocity)
sel = !(as.vector(dists) == 0)
plot(as.vector(corv[sel]) ~ as.vector(dists[sel]),
	xlim = c(0,500), ylim = c(.4, 1), xlab = "distance (km.)", 
	ylab = "correlation") 
# plots all points twice, ignores zero distance 
```

```{r}
# now really get fig 3:
ros = rownames(corv) == "ROS"
dists.nr = dists[!ros,!ros]
corv.nr = corv[!ros,!ros]
sel = !(as.vector(dists.nr) == 0)
plot(as.vector(corv.nr[sel]) ~ as.vector(dists.nr[sel]), pch = 3,
	xlim = c(0,500), ylim = c(.4, 1), xlab = "distance (km.)", 
	ylab = "correlation") 
# add outlier:
points(corv[ros,!ros] ~ dists[ros,!ros], pch=16, cex=.5)
xdiscr = 1:500
# add correlation model:
lines(xdiscr, .968 * exp(- .00134 * xdiscr))
```

# demo(wind)
```{r}
# PLEASE read the vignette of package spacetime for a more
# clever way to do all this!
library(sp)
library(rgdal)
library(maptools)

m = map2SpatialLines(
	map("worldHires", xlim = c(-11,-5.4), ylim = c(51,55.5), plot=F))
proj4string(m) = "+proj=longlat +datum=WGS84 +ellps=WGS84"
m = spTransform(m, CRS("+proj=utm +zone=29 +datum=WGS84 +ellps=WGS84"))

# model temporal autocorrelation
acf(wind[7])
tdiscr = 0:40
lines(tdiscr, exp(- tdiscr/1.5))
```

## set up data, last year
```{r}
years = 61
months = 1
jday = c(1,6,11,16,21,26)
sel = wind[wind$year %in% years & 
	wind$month %in% months &
	wind$jday %in% jday,]

#stations = 4:15
stations = 4:15

sels = stack(sel[stations])
sels$t = rep(sel$jday, length(stations))
sels$x = coordinates(wind.loc)[match(sels$ind, wind.loc$Code),1]
sels$y = coordinates(wind.loc)[match(sels$ind, wind.loc$Code),2]
summary(sels)

coordinates(sels) = ~x+y
proj4string(sels) = "+proj=longlat +datum=WGS84 +ellps=WGS84"
sels = spTransform(sels, CRS("+proj=utm +zone=29 +datum=WGS84 +ellps=WGS84"))
grd = makegrid(m, n = 1000)
grd$t = rep(1, nrow(grd))
coordinates(grd) = ~x1+x2
gridded(grd)=TRUE
proj4string(grd) = proj4string(sels)

#sels = as(sels, "data.frame")

# setup grid
covfn = function(x, y = x) { 
	u = spDists(coordinates(x), coordinates(y))
	t = abs(outer(x$t,y$t,"-"))
	0.6 * exp(-u/750000) * exp(-t/1.5)
}
for (i in 1:120) {
	grd$t = rep(i/4, nrow(grd))
	n = paste("t", i/4, sep="")
	grd[[n]] = krige0(sqrt(values)~1, sels, grd, covfn)
}
grd$t = NULL
#grd$pr = out$pred
#library(lattice)
#levelplot(pr~x1+x2|t,grd,col.regions=bpy.colors())
spl = list(list("sp.points", sels,first=F, cex=.5),
	list("sp.lines", m, col='grey'))
spplot(grd, sp.layout = spl, col.regions=bpy.colors())
```

#  The other approach, using spacetime classes:
```{r}
```{r}
data("wind", package = "gstat")
wind[1:6,1:12]

library(spacetime)


library(xts)
.parseISO8601('2010-05')
.parseISO8601('2010-05-01T13:30/2010-05-01T13:39')
```

```{R}
library(gstat)
library(spacetime)
library(RColorBrewer)
data("wind")
wind.loc$y = as.numeric(char2dms(as.character(wind.loc[["Latitude"]])))
wind.loc$x = as.numeric(char2dms(as.character(wind.loc[["Longitude"]])))
coordinates(wind.loc) = ~x+y
proj4string(wind.loc) = "+proj=longlat +datum=WGS84"

library(mapdata)
plot(wind.loc, xlim = c(-11,-5.4), ylim = c(51,55.5), axes=T, col="red",
	cex.axis =.7)
map("worldHires", add=TRUE, col = grey(.5))
text(coordinates(wind.loc), pos=1, label=wind.loc$Station, cex=.7)

wind[1:3,]
wind$time = ISOdate(wind$year+1900, wind$month, wind$day)
wind$jday = as.numeric(format(wind$time, '%j'))


stations = 4:15
windsqrt = sqrt(0.5148 * as.matrix(wind[stations])) # knots -> m/s
Jday = 1:366
windsqrt = windsqrt - mean(windsqrt)
daymeans = sapply(split(windsqrt, wind$jday), mean)
meanwind = lowess(daymeans ~ Jday, f = 0.1)$y[wind$jday]
velocities = apply(windsqrt, 2, function(x) { x - meanwind })


wind.loc = wind.loc[match(names(wind[4:15]), wind.loc$Code),]
pts = coordinates(wind.loc[match(names(wind[4:15]), wind.loc$Code),])
rownames(pts) = wind.loc$Station
pts = SpatialPoints(pts, CRS("+proj=longlat +datum=WGS84 +ellps=WGS84"))

library(rgdal)
utm29 = CRS("+proj=utm +zone=29 +datum=WGS84 +ellps=WGS84")
pts = spTransform(pts, utm29)

wind.data = stConstruct(velocities, space = list(values = 1:ncol(velocities)), 
	time = wind$time, SpatialObj = pts, interval = TRUE)
class(wind.data)

library(maptools)
m = map2SpatialLines(
	#map("worldHires", xlim = c(-11,-5.4), ylim = c(51,55.5), plot=F))
	map("worldHires", xlim = c(-11.5,-6.0), ylim = c(51.3,55.0), plot=F))
proj4string(m) = "+proj=longlat +datum=WGS84"
m = spTransform(m, utm29)

grd = SpatialPixels(SpatialPoints(makegrid(m, n = 300)),
	proj4string = proj4string(m))

wind.data = wind.data[, "1961-04"]

n = 10
library(xts)
tgrd = seq(min(index(wind.data)), max(index(wind.data)), length=n)
pred.grd = STF(grd, tgrd)

v = vgmST("separable", space = vgm(1, "Exp", 750000), time = vgm(1, "Exp", 1.5 * 3600 * 24),
         sill=0.6)
wind.ST = krigeST(values ~ 1, wind.data, pred.grd, v)
colnames(wind.ST@data) <- "sqrt_speed"

layout = list(list("sp.lines", m, col='grey'),
  list("sp.points", pts, first=F, cex=.5))
stplot(wind.ST, col.regions=brewer.pal(11, "RdBu")[-c(10,11)],
  at=seq(-1.375,1,by=.25),
  par.strip.text = list(cex=.7), sp.layout = layout)

library(lattice)
library(RColorBrewer)
b = brewer.pal(12,"Set3")
par.settings = list(superpose.symbol = list(col = b, fill = b), 
	superpose.line = list(col = b),
	fontsize = list(text=9)) 
print(stplot(wind.data, mode = "ts",  auto.key=list(space="right"), 
	xlab = "1961", ylab = expression(sqrt(speed)),
	par.settings = par.settings))


scales=list(x=list(rot=45))
stplot(wind.data, mode = "xt", scales = scales, xlab = NULL, 
  col.regions=brewer.pal(11, "RdBu"),at = seq(-1.625,1.125,by=.25))
```

## Empirical orthogonal functions:
```{r}
eof.data = eof(wind.data)
eof.int = eof(wind.ST)
eof.xts = eof(wind.ST, "temporal")
spplot(eof.int[1:4], col.regions=bpy.colors(),
	par.strip.text = list(cex=.5), as.table = TRUE, sp.layout = layout)
```

Further reading: Spatio-Temporal Interpolation using gstat PDF download
Benedikt Gräler, Edzer Pebesma and Gerard Heuvelink , The R Journal (2016) 8:1, pages 204-218; https://journal.r-project.org/archive/2016/RJ-2016-014/index.html

